---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# eList

<!-- badges: start -->
<!-- badges: end -->

Write list comprehensions in R! 

The eList package allows users to write vectorized `for` loops and contains a variety
of tools for working with lists and other vectors. Just wrap a normal `for` loop 
within one of the comprehension functions, such as `List()`, and let the package 
do the rest.

Features include, but are not limited to:

* Multiple variables within the loop using `"."` to separate names.
* Multiple returns types: list, numeric vectors, character vectors, matrices, environments, etc.
* `if`, `else` statements to filter results.
* Name assignment within the loop using `=`.
* Parallelization by adding the option `, clust = ` and the cluster.
* Nested loops.
* Helper functions such as `enum` and `items` to access the index or name within the loop, or `zip` objects together.
* Summary comprehensions that calculate sums, means, etc. from `for` loops.
* Use either `List(...)` or `List[...]` for the comprehension. 
* Higher order functions, such as map or filter, that accept functions built from formulas *(using `.` notation for variables)* or calls.


## Installation

You can install the released version of eList from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("eList")
```

## Examples

A simple "list" comprehension that accumulates all integer sequences to 4 using the `List` function. Though it looks like a `for` loop, it is actually using `lapply` behind the scenes. 

```{r example}
library(eList)
List(for (i in 1:4) 1:i)
```

Loops can be nested and filtered using `if` statements. The example below uses `Num` to produce a numeric vector rather than a list. Other comprehensions include `Chr` for character vectors, `Logical` for logical vectors, `Vec` for flat *(non-list)* vectors, etc.

```{r}
Num(for (i in 1:4) for (j in 3:6) if (i == j) i^2)
```
Use the "dot" notation to use multiple variables within the loop. 

```{r}
Chr(for (i.j in zip(1:4, 5:8)) paste0(i, j))
```
Use `=` within the loop to assign a name to each item within the list, or other item.

```{r}
values <- zip(letters[1:4], 5:8)

List(for (i.j in values) i = j)
```

Parallelization is also very easy. Just create a cluster and add it to the comprehension with the `clust` argument.

```{r}
my_cluster <- auto_cluster()

x <- Num(for (i in sample(1:100, 50)) sqrt(i), clust = my_cluster)

# Close the cluster if not needed!
close_cluster(my_cluster)

x
```

Want a statistical summary using a comprehension? eList contains a variety of summary functions for that purpose. `Stats` is a general summary comprehension that computes many different values. 

```{r}
Stats(for (i in sample(1:100, 50)) sqrt(i))
```

eList also contains functional programming style functions for working with lists and other vectors. These functions perform an operation using a function on another object. They are similar to the higher order functions in Base R, but are pipe-friendly, handle a wide ranger of object types, and allow for different methods of specifying functions.

```{r}
x <- list(1:4, 5:8, 9:12)
map(x, mean)
```

This can also be calculated using formula notation. Formulas can be be written as either a two-sided formula or a one-sided formula by prefixing variables with dots. 

```{r}
# Two-sided Formula
map(x, i ~ sqrt(i) + 1)

# One-sided Formula
map(x, ~ sqrt(.i) + 1)
```

The higher order functions also accept unevaluated "calls".

```{r}
round2 <- substitute(round(digits=2))
map(rnorm(5), round2)
```

